Investigating Determinism of Physics Engines for usage in CAV and V\&V simulations

### 2022-01-24 document/source/body.tex
shortened FP and scheduling

A common misconception is to attribute non-determinism to the use of floating point number representation. This is an erroneous conflation with non-deterministic computational execution.  
%
Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. 
Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value. This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}. %

As a consequence, floating-point arithmetic is not associative. Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}. This is because rounding is being performed at different stages in the computation, depending on the respective execution order. 
%
% Translating to the usage of AV simulations in physics engines,
In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions.
%
Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability. In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.

In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. 
%
It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. 
%
Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. 
%
So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.

% However, as illustrated above on the example of non-associative addition, a change of execution order can lead to sequences of floating-point operations producing different results. Thus, any uncontrolled options to modify execution order, e.g.\ at the compiler level, within the runtime environment or at the hardware level, can cause simulation results to differ between runs, resulting in non-zero variance and loss of determinism.

% Beyond that, aggressive optimisations by the compiler can also introduce incorrectness in floating-point arithmetic~\cite{llvm-floating-point}. However, the same executable should still return the same output for identical input.

% Execution order changes can occur due to the use of a different compiler or as a consequence of parallelisation, e.g.\ within the runtime environment or at the hardware level. 
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, and 
parallelisation within the runtime environment or at the hardware level can change the execution order, which can cause simulation results to differ between runs when floating-point operations are affected. 
%
Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.

In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.

\subsection{Scheduling, Concurrency and Parallelisation}
Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. 
%
In principle, changing the scheduling policy and thread priorities may increase simulation variance.

It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration. Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.

However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events. This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance.
%
Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}.
%

Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.

Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes. This may be decided based on factors such as throughput, latency or CPU utilisation. 
%
If due to the CPU utilisation policy the same single-threaded script executes multiple times across different physical cores of the same CPU type,  then execution should still produce the same output. 
%
This is because the processor cores are identical and any impurities across the bulk silicon and minor perturbations in the semiconductor processing across the chip that may exist should have been accommodated for in design and manufacturing tolerances.
%
However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.

Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.
%

Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously. While this would be faster than a serial execution, the order of execution arising from program-level concurrency is often not guaranteed.

@@

@@

\subsection{Actor Navigation}
% There is evidence to suggest actor navigation, typically pedestrians, could be a source of non-deterministic simulation behaviour~\cite{CARLABenchmark}. Unreal Engine is the underlying framework for the CARLA simulator which uses the A* search algorithm to find optimal paths for actor navigation~\cite{newton2016unreal}. 
%
In game engines, the A* search algorithm is often used for navigation to find optimal paths for actors. 
%
The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.

@@

% The A* algorithm~\cite{AStarBook} will give deterministic outcomes as long as the environment is deterministic~\cite{AirsimUnrealArticle, UnrealAIDocumentation}. A navigation mesh is a fixed area of the environment where actors are free to navigate. Navigation meshes are used by the path planning algorithm to find the shortest distance between navigable points in then environment and could be considered as a potential source of non-determinism if the navigation mesh is not a fixed entity. However, the environment management in CARLA implements fixed binary files for navigation meshes that are linked to each road scene or driving environment. These cannot be unknowingly modified and therefore can be considered fixed.

% Unless other sources that can alter the fixed environment exist, the A* algorithm should give deterministic results. Thus, actor navigation should not be considered a source of non-determinism.

>>>>

% NEW TEXT V3
It is a common misconception to attribute non-deterministic computational execution with the use of floating point number representation, which limited by the fixed bit width available and potentially resulting rouding errors~\cite{FloatingPointsBook,goldberg1991every}. As a consequence, floating-point arithmetic is not associative~\cite{Kapre2007} and depends on execution order where some authors even suggest avoiding floating point representation entirely~\cite{empirical-analysis-of-flaky-tests}. 
%
In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions but we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. So, even if the result of a operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}. 
%
However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, 
parallelisation within the runtime environment or at the hardware level and if the execution is performed on a GPU rather than a CPU (which may have different register widths~\cite{Whitehead2011}) may all affect the execution order. 
%
In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable, hardware, configuration and execution order.

% PREVIOUS TEXT V2
% A common misconception is to attribute non-determinism to the use of floating point number representation. This is an erroneous conflation with non-deterministic computational execution. Floating-point representation of real numbers is limited by the fixed bit width available in the hardware, resulting in the finite precision with which numbers can be represented in a computation. Thus, the results of arithmetic operations must fit into the given bit width, which is achieved through rounding to the nearest representable value. This gives rise to rounding errors~\cite{FloatingPointsBook,goldberg1991every}.

% As a consequence, floating-point arithmetic is not associative. Thus, for arithmetic expressions beyond two operands, such as $x+y+z$, executing $x+(y+z)$ rather than $(x+y)+z$, can produce different results~\cite{Kapre2007}. This is because rounding is being performed at different stages in the computation, depending on the respective execution order. In the context of AV simulation, such rounding errors could result in accuracy issues of, for example, actor positions within the environment, leading to falsely satisfied or failed assertions. Thus, one could conclude that floating-point arithmetic causes non-determinism, resulting in loss of repeatability. In fact, some authors suggest avoiding the use of floating-point representation entirely for assertion testing due to imprecision and non-deterministic behaviour~\cite{empirical-analysis-of-flaky-tests}.

% In contrast, we argue that the precision issues related to floating-point operations are better described as \textit{incorrectness} that is in fact repeatable; they do not cause non-determinism per se. It is normally reasonable to assume that the processors on which game engines run are deterministic with respect to floating-point arithmetic. Thus, given the same operands and sequence of operations, the same processor should always produce the same bit pattern. So, even if the result of a floating-point operation is incorrect due to floating-point rounding errors, it should always be equally \textit{incorrect} for implementations that conform to the IEEE floating-point standard~\cite{8766229}.

% However, different compiler configurations, aggressive optimisations~\cite{llvm-floating-point}, and 
% parallelisation within the runtime environment or at the hardware level can change the execution order, which can cause simulation results to differ between runs when floating-point operations are affected. Furthermore, floating-point calculation results may differ if the execution is performed on a GPU rather than a CPU which may have different register widths~\cite{Whitehead2011}.

% In conclusion, floating-point arithmetic does not cause non-zero \textit{simulation variance} for repeated simulation runs when using the same executable on the same hardware with exactly the same configuration and execution order.

\subsection{Scheduling, Concurrency and Parallelisation}

Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority depending on the operating system's scheduler policy. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. In principle, changing the scheduling policy and thread priorities may increase simulation variance.

It is unlikely that changes to thread priorities or scheduling policy would occur during repeated tests on the same hardware and should give the same output. However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading} or if the scheduler simply randomly selects from a set of threads with equal priority, this may alter an otherwise deterministic sequence of events.

Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes. This may be decided based on factors such as throughput, latency or CPU utilisation. Scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, or if executing across parallel GPU architecture, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution. Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.

@@

% PREVIOUS TEXT V2
% Runtime scheduling is a resource management method for sharing computational resources between tasks of different or equal priority where tasks are executed depending on the operating system's scheduler policy. A scheduler policy may be optimised in many ways such as for task throughput, deadline delivery or minimum latency~\cite{liu1973scheduling}. In principle, changing the scheduling policy and thread priorities may increase simulation variance.

% It is, however, unlikely that changes to thread priorities or scheduling policy would occur during repeated controlled tests for the same hardware and operating system configuration. Given that hardware is considered deterministic and if the thread execution order does not change between tests, then for the same hardware and operating system configuration the same output should be given.

% However, if some aspects of the game loop are multi-threaded~\cite{unity_multithreading}, then, even with a clear thread scheduling order, any background process may interrupt the otherwise deterministic sequence of events. This may, for example, alter the number of physics calculations that can be performed within the game loop and hence result in simulation variance. Using multiple threads has been found to affect initialisation ordering for training machine learning models which can lead to unpredictable ordering of training data and non-deterministic behaviour~\cite{Sculley2015,Breck2017}. Interference may also occur when a scheduler simply randomly selects from a set of threads with equal priority, resulting in variation of the thread execution order.

% Similar to thread scheduling, scheduling at the hardware level on a multi-core system determines on which processor core to execute processes. This may be decided based on factors such as throughput, latency or CPU utilisation. If due to the CPU utilisation policy the same single-threaded script executes multiple times across different physical cores of the same CPU type,  then execution should still produce the same output. This is because the processor cores are identical and any impurities across the bulk silicon and minor perturbations in the semiconductor processing across the chip that may exist should have been accommodated for in design and manufacturing tolerances. However, scheduling multiple processes across several processing cores, where the number of cores is smaller than the number of processes, can result in variation of the execution order and cause simulation variance unless explicitly constrained or statically allocated prior to execution.

% Indeed, the developers of the debugging program \texttt{rr}~\cite{RR_link} took significant steps to ensure deterministic behaviour of their program by executing or context-switching all processes to a single core, which avoids data races as single threads cannot concurrently access shared memory. This allowed control over the scheduling and execution order of threads, thus promoting deterministic behaviour by design~\cite{acm-q-rr-interview}.

% Likewise, simulation variance may be observed for game engines that use GPU parallelisation to improve performance by offloading time-critical calculations to several dedicated computing resources simultaneously. While this would be faster than a serial execution, the order of execution arising from program-level concurrency is often not guaranteed.

% Overall, scheduling, concurrency and parallelisation may be reasons for \textit{simulation variance}.

@@

\subsection{Actor Navigation}
In game engines, the A* search algorithm is often used for navigation to find optimal paths for actors. 
%
The optimality and validity of A* solutions is only guaranteed if the environment is deterministic, and if not, then there are no guarantees that an execution of the same A* solution will lead to the same outcome across multiple runs.

@@
